---
title: "Final Project Proposal"
subtitle: "Team 1"
format: html
editor: visual
---

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)
```

# Data 1

## Introduction and data

-   The source of the data is data.medicaid.gov.

-   It was originally collected in 2018 and is collected by reports of each state for covered outpatient drugs that are paid for by state Medicaid agencies since the start of the Medicaid Drug Rebate Program.

-   I can see that there are a wide range of states, if not all of the states listed for the usage of a large sum of drugs, however i will only be focusing on Prozac which is a commonly used anti-depressant drug.

## Research question

-   What is the utilization of Prozac like in various states for outpatients that have medicaid in 2013 and why was the utilization so high or low, what factors impacted the utilization rates?

-   The research topic has a lot to do with trying to tie the rate of prozac usage to depression levels of medicaid users in 2013. This then allows us to view the outside factors that might have been impacting depression levels in 2013. If prozac suppression usage is high, then in the year 2013 depression levels were high as well especially in black and brown communities because they are the majority of medicaid users.

-   There are both categorical and quantitative variables, the categorical consist of the type of drug, suppression usage(true or false), and states. The quantitative is package sizes, year and quarters.

## Literature

-   https://www.ncbi.nlm.nih.gov/books/NBK459223/

-   This article goes into the depth the impact of Prozac in various situations as well as why Prozac is used and for what kinds of patients. It outlines the FDA approved portions of the drug and the parts that are left to therapist and mental health specialist. Furthermore, it gives details of the adverse effects and symptoms of the drug. What this article does more than any others is gives advice for enhancing healthcare team outcomes and places preganant women and children at the forefront because children as young as 8 can use Prozac.

-   My research question builds on and is different from the article because its focus on the usage of prozac in different states in a particular year and allows space to dive into the political and social sphere of that specific year. Rather the article is more generalized and not specific to state usage numbers, medicaid and one particular year. \## Glimpse of data

```{r}
#| label: load-data-1

prozac_utilization <- read_csv("data/data.csv")

glimpse(prozac_utilization)
```

# Data 2

## Introduction and data

-   The source of the data is from CORGIS: The Collection of Really Great, Interesting, Situated Datasets, originally stored in the Substance Abuse & Mental Health Data Archive (SAMHDA).

-   The data in the data set was collected from individual states from the National Survey on Drug Use and Health, which is taken once a year. The data ranges from 2002 to 2018.

-   From the data, we can see that every state in the United States is accounted for each year. The different drugs recorded in the data set are alcohol, tobacco, cocaine, and marijuana as well as an age range for each of them. There are 3 age ranges from 12-17, 18-25, and 26+. Also, there are totals and rates for each of the drugs that are recorded in the data set.

## Research question

-   Do people aged 17-25 abuse tobacco or marijuana more on average in 2018? What about in 2002?

-   From this question, I aim to see if marijuana usage is increasing and tobacco use is decreasing in adolescents and young adults over the years. From here, we can uncover possible reasons for the data we find.

-   There are both categorical and quantitative variables involved in my research question. The states in the data set are categorical while the totals and rates of tobacco and marijuana are quantitative.

## Literature

-   Find one published credible article on the topic you are interested in researching.

    -   https://infoweb-newsbank-com.proxy.lib.duke.edu/apps/news/document-view?p=WORLDNEWS&t=&sort=\_rank\_%3AD&fld-base-0=alltext&maxresults=20&val-base-0=are%20adolescents%20beginning%20to%20abuse%20marijuana%20more%20than%20tobacco%20&docref=news/16E335BBF826EC78

-   Provide a one paragraph summary about the article.

    -   This article explores the usage of vapes that contain THC. They look at the change in popularity from cigarettes (tobacco) to e-cigarettes and vaping (nicotine) to new vapes that contain marijuana substances. Now, these vapes are becoming increasingly easy to access, leading to the rise of adolescents and young adults being exposed to marijuana.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

    -   My research question will explore the rise or fall in marijuana in adolescents and young adults, building on the article cited.

## Glimpse of data

```{r}
#| label: load-data-2

drugs <- read_csv("data/drugs.csv")

glimpse(drugs)
```

# Data 3

## Introduction and data

-   Identify the source of the data.

<!-- -->

    -   The jail and jail death data were found at reuters.com.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

    -   In 2020, Reuters journalists filed over 1500 public records requests to gain death data from 2008 to 2019 in the nation's biggest jails. The data came form the Justice Department, who collected jail death.

-   Write a brief description of the observations.

    -   The majority of the inmates who died were never convicted of charges.

    -   Over 80% of the deaths in the jails were of male inmates.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

    -   What does the relationship between factors such as race, custody status, jail size, and year affected the of deaths in the 10 largest jails in each state?

    -   Of the unconvicted inmates who died, what were the most common reasons of death?

        -   Maybe looking specifically at suicides

    -   Does the type of medical provider affect the death rate in each jail?

        -   Maybe look at freqs of each reason grouping by type of medical provider

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

    -   With the Justice Department hiding the jail death data, this research topic hopes to look into the injustices regarding crimes and jails, where certain populations are disproportionately dying in the jails. Especially considering how many inmates die waiting for their trial, this topic hopes to dive deeper into some of the trends found in the data and potentially find populations that are disproportionately affected.

    -   Looking at the inmate death data, we believe that unconvicted Black men will be disproportionately represented among the deaths in most of the largest 10 jails in each state.

-   Identify the types of variables in your research question. Categorical? Quantitative?

    -   Race, gender, jail, and custody_status are all categorical variables.

## Literature

-   Find one published credible article on the topic you are interested in researching.

    -   https://www.nytimes.com/2022/11/22/us/jails-deaths.html

-   Provide a one paragraph summary about the article.

    -   Dewan explores the reasons of why the death rate of inmates in jails have been growing over the past years, many of which can be attributed to the dangerous conditions in the jails. For example, sometimes inmates were denied meals and bathroom breaks, and jails fail to properly treat or accomodae inmates' severe mental illnesses and disabilities. Also, it was reported that the Justice Department at one point "recorded only 39 percent of deaths in local jails," demonstrating how the terrible conditions and the severity of the issue becomes hidden, which makes it harder for policymakers and activists to become aware or speak out about these issues. Especially when about 2/3 of the inmates who die are not convicted, it makes the situation worse that the jails don't have human decency or provide basic needs for inmates.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

    -   The research question goes more into details of different jails or states and how race and gender is represented in the deaths in the data.

## Glimpse of data

```{r}
#| label: load-data-3

jail_deaths <- read_csv("data/all_deaths.csv")

glimpse(jail_deaths)

jails <- read_csv("data/all_jails.csv")

glimpse(jails)
```
